---
title: "regular expressions"
output: html_notebook
---

Regular expressions are what we use for pattern matching in text e.g. rather than looking for "Edinburgh", looking for a place name - with a capital letter followed by lower case

Used for things like - is this email address valid? Needs text, then @ sign, ending with something .com/co.uk. Expects certain character and issues warning when not seen

```{r}
single_string <- "string a"
strings <- c("string a", "string b", "string c", "string d", "striiiiiiing e")
```

```{r}
library(tidyverse)

pattern <- "a" # gonna look for this character a
```

Is this pattern present in a string?
--> str_detect()

```{r}
single_string %>% 
  str_detect(pattern, negate = TRUE)
```
Returns TRUE as 'a' is in 'string a'
If we set negate to TRUE (default FALSE), we just flip the search "does it not appear here?"

```{r}
strings %>% 
  str_detect(pattern)
```
Can check if 'a' is any of the strings in the multi object

```{r}
pattern <- "str"

strings %>% 
  str_detect(pattern)
```

5 trues returned as each string has str in
If the pattern was 'string' then the final one would fail as it's striiiiiing

Match for multiple characters, not in order just at least one of them
```{r}
pattern <- "[abc]" # is any one character in the string an a, b or c?
strings %>% 
  str_detect(pattern)
```
string a yes, string b yes, string c yes, but d and e no

```{r}
pattern <- "[a-c]" # what about any letters between a-c?
strings %>% 
  str_detect(pattern)
```

```{r}
pattern <- "[a-z]" # within the range of the alphabet
strings %>% 
  str_detect(pattern)
```

Everything matches as every string has at least one character in the alphabet

What about with capitals?
```{r}
pattern <- "[A-Z]" # upper case letters
strings %>% 
  str_detect(pattern)
```

All false as function is case-sensitive

could do "[a-zA-Z]" to cover all bases
or "[A-z]" as tidyverse stores upper alphabet then lower


```{r}
pattern <- "[A-z]" # upper or lower
strings %>% 
  str_detect(pattern)
```

(?i) - "mode modifier" to make it case insensitive
```{r}
pattern <- "(?i)[A-Z]" 

strings %>% 
  str_detect(pattern)
```

Aside: dates
Can slowly build up the pattern we're looking for with like [19-20][0-9][0-9][01:12][.,-/] sort of thing

What if we want 3 consective i's?
```{r}
pattern <- "iii"
```
What about 50 i's? Don't want to type them all

```{r}
pattern <- "i{50}"

string <- "hiiiiiiiiiiiiiiiiiiiiiiiiiiiiiiiiiiiiiiiiiiiiiiiiiiiiii"

string %>% 
  str_detect(pattern)
```


Task - 15 mins
Make a vector containing the following strings:
"We're having great fun learning about data analysis at CodeClan!"
"CodeClan is at 37 Castle Terrace in Edinburgh and ClockWise in Glasgow"
"Our cohort code starts with DE or DR"
"You get a great view of the castle or the national conservatoire from just outside CodeClan's door"

```{r}
strings <- c(
  "We're having great fun learning about data analysis at CodeClan!",
"CodeClan is at 37 Castle Terrace in Edinburgh and ClockWise in Glasgow",
"Our cohort code starts with DE or DR",
"You get a great view of the castle or the national conservatoire from just outside CodeClan's door")
```

Write and test regular expressions which will match:
Any string mentioning CodeClan


```{r}
pattern <- "CodeClan"

strings %>% 
  str_detect(pattern)
```

Any string containing two consecutive capital letters
```{r}
pattern <- "[A-Z]{2}"

strings %>% 
  str_detect(pattern)
```
Any string not including the word "castle" or "Castle"
Hint
Remember, str_detect() can take an optional third argument called negate
Hint
The techniques we looked at above can be combined.

```{r}
pattern <- "(?i)castle" 
pattern <- "[cC]astle" # or this sexy little method

strings %>% 
  str_detect(pattern, negate = TRUE)
```

We've covered DO they exist, but what about extracting them

# Extracting Matching Substrings

```{r}
strings <- c(
  "This string has an_address@email.com in it", 
  "This one has user.name@company.co.uk", 
  "Now we've got other_person_123@server.net and my.name@runningoutofideas.com"
)
```

`str_extract()` - a cousing of str_extract

```{r}
pattern <- "[a-z]@" # a lteer before an @ sign

strings %>% 
  str_extract(pattern)

```

Common quantifiers:

*: 0 or more
+: 1 or more
{x}: exactly x

```{r}
pattern <- "[a-z]*@" # any amount (0+) of letters (*) before an @ sign
pattern <- "[a-z]+@" # 1+ letters (*) before an @ sign

strings %>% 
  str_extract(pattern)

```

Doing well here, it's getting the first halves of the addresses

```{r}
# pattern <- "[a-z0-9]+@" # numbers or letters
pattern <- "[:alnum:]+@[:lower:]+" # can also use alnum

strings %>% 
  str_extract(pattern)

```

In string 3, only the first email address is extracted (which has numbers), want it to extract both

```{r}
pattern <- "[a-z0-9_\\.]+@[a-z\\.]+" # include underscores and full stops
# include double backslashes so that we get period, not its special regex meaning
strings %>% 
  str_extract(pattern)

```

Got 3 out of 4 email addresses now
Use `str_extract_all()` to get all matching cases, not just the first in each string

```{r}
pattern <- "[a-z0-9_\\.]+@[a-z\\.]+" # include underscores and full stops
# include double backslashes so that we get period, not its special regex meaning
strings %>% 
  str_extract_all(pattern)

```
Note the extra layer of hierarchy: a list (1, 1, 2) denoted by double square brackets [[]]

Note how we can write regex in normal strings, no fancy packages needed. Is a common language. Hence why we need double backslashes to negate the other meanings of characters

```{r}
pattern <- "\\S*@\\S*" # here is a bit of a cheaty way, using the S not white space character class

# however this would work for "adf@afaf" so isn't very selective in general use
strings %>% 
  str_extract_all(pattern)

```

str_replace() allows us to replace chunks of strings

```{r}
pattern <- "[a-z0-9_\\.]+@[a-z\\.]+"  

strings %>% 
  str_replace_all(pattern, "REDACTED")

```


Extra stuff

```{r}
punct_strings <- c(
  "Why is this a question?", 
  "I am a sentence with no punctation",
  "I am a well punctuated sentence.", 
  "WOW!"
)
```
Extract punctuation
```{r}
punct_strings %>% 
  str_extract_all("[:punct:]")
```

Extract 3 letter word followed by punctuation
```{r}
punct_strings %>% 
  str_extract_all("[A-Z]{3}[:punct:]")
```

Anchors
^: start
$: end

```{r}
test_strings <- c(
  "abc", 
  "Abc",
  "abC",
  "s"
)
```

Below extracts any letters
```{r}
test_strings %>% 
  str_extract_all("[a-z]")
```

What if we want the pattern to have to match the start of the string
```{r}
test_strings %>% 
  str_extract_all("^[a-z]") # has to start with a lower case

test_strings %>% 
  str_extract_all("[a-z]$") # has to finish with a lower case

test_strings %>% 
  str_extract_all("^[a-z]+$") # has to start and end with a lower case, need the plus to specify more than 1
```


Strings within a tibble

Set up a tibble:
```{r}
student <- c("Abigail", "Bjorn", "Caitlin")
age <- c(25, 56, 40)
emails <- c("abigal12@company.co.uk", "bsmith@gmail.co.uk", "caitlin_1980")

student_info <- tibble(student, age, emails)

student_info 
```

```{r}
student_info %>% 
  str_detect(emails, "[a-z]+@") # too many arguments into function --> error

student_info %>% # needs to have dplyr function
  mutate(email_detect = str_detect(emails, "[a-z]+@"))


```
Return boolean depending on if email address is valid
```{r}
student_info %>% 
  mutate(emails = str_detect(emails, "@[a-z\\.]+"))
```

<rebus> package is cool (optional)
Unicode can be used for writing regex expressions (uncode)


