---
title: "Recoding exercise"
output: html_notebook
---

Outcomes
- some additional functions:
-- distinct()
-- count()

-- if_else()
-- recode()
-- case_when()

```{r}
library(tidyverse)
library(CodeClanData)
```

`distinct()` is a dplyr function that takes in a dataset and a list of columns, and returns the unique values in that column

`count()` is another dplyr function that returns the number of occurrences of each distinct level in a column

```{r}
students_big %>% 
  distinct(region)
# shows us the distinct regions, no copies
# how many times does e.g. USA occur? use count
students_big %>% 
  count(region)
# 2 people

# can do multiple columns
students_big %>% 
  count(region, gender, sort = TRUE) # sort true orders from most prevalent to least
```

Why would we need to recode?
What is recoding?

It's a large part of data cleaning - involves changing or recoding the values in columns

e.g. Edinburgh vs Edinburgh, Scotland
e.g. EH91AA vs EH9 1AA

Reduce level of granularity
e.g. Country: Scotland, England, France, Wales
--> Country: UK, UK, France, UK

Three methods for us:
`if_else()`
`recode()`
`case_when()`
Each performs recode in its own way

## if_else()

```{r}
letters <-  c("a", "B", "c")

if_else(letters == "B", "b", "e")
# replace upper case B with lower case b, everything else with e. FUnction needs the 2nd argument (e) to work

if_else(letters == "B", "b", NA_character_)
# replaces with NA
if_else(letters == "B", "b", letters)
# if condition not met, just left alone

```

what if we want male or female in gender col
i.e. replace M with Male, F with Female
```{r}
students_big %>% 
  mutate(gender = if_else(gender == "M", "Male", "Female")) %>% 
  count(gender)
# count tells us that we've covered all options, there's nothing that falls outside of male and female
```


Task - 5 minutes

1. Change the column handed to to "Right" and "Left" instead or "R" and "L".
2. In favourite_physical_activity change "Football/Soccer" into "Football" and leave all other activities the same.

```{r}
students_big %>% 
  mutate(handed = if_else(handed == "R", "Right", "Left")) %>% 
  count(handed)
# careful here as some students have B for both, so could do 2 if_elses for right and left, to make sure Bs are left alone, or 3 if_else's to cover all bases and call B both

students_big %>% 
  mutate(favourite_physical_activity = if_else(favourite_physical_activity == "Football/Soccer", "Football", favourite_physical_activity)) %>% 
  distinct(favourite_physical_activity)

```

# `recode()`

recode(column/vector, "old_value" = "new_value", "other_old_value" = "other_new_value", ..., .default = NULL # if it's not in your replacement pairs, can set a default replacement #, .missing = NULL)

```{r}
letters

# recode letters so that they are all upper case
recode(letters, "a" = "A", "c" = "C")

# recode handed so that L -> left, R -> right, B -> both
students_big %>% 
  select(handed) %>% 
  mutate(handed = recode(handed, "L" = "Left", "R" = "Right", "B" = "Both")) %>% 
  count(handed)


# recode the travel to school column such that

# Car -> car
# Bus -> bus
# Walk -> walk
# Anything else -> other

students_big %>% 
  mutate(travel_to_school = recode(travel_to_school,
                                   "Car" = "car",
                                   "Bus" = "bus",
                                   "Walk" = "walk",
                                   .default = "other")) %>% 
  count(travel_to_school)

# if we haven't explicitly stated it, then make it 'other'
```
```{r}
data <- read.table(header = TRUE, text="
 subject sex control cond1 cond2 compl
       1   M     7.9  12.3  10.7  T
       2   F     6.3  10.6  11.1  T
       3   F     9.5  13.1  13.8  N
       4   M    11.5  13.4  12.9  F
")
```

Task - 7 minutes
Using the data below, do the following:
1. Recode the sex column to be “Male” and “Female”
2. Recode the compl column to be T = “True”, F = “False”, and NA otherwise

```{r}
data %>% 
  mutate(sex = recode(sex,"M" = "Male",
                      "F" = "Female")) %>% 
  mutate(compl = recode(compl, "T" = "True",
                         "F" = "False",
                         .default = NA_character_))

# can also do both mutates in one
data %>% 
  mutate(sex = recode(sex,"M" = "Male",
                      "F" = "Female"),
         compl = recode(compl, "T" = "True",
                         "F" = "False",
                         .default = NA_character_))
# have to use NA_character_ as NA is logical type
```


# `case_when()`

The most adaptable of the 3 but more complex

A bit like multiple if_else()s
if a condition -> do this
if a different condition -> do this
if a different condition -> do this
if a different condition -> do this
etc

case_when(
LHS conditional test ~ value when test is met
)

use tilda ~ to separate idea of LHS and RHS

with case_when, the order we define our conditional expressions matters. If a condition is met (evaluate to TRUE), case_when doesn't check the later expressions

```{r}
letters

# using case_when, change all of the letters to upper case
case_when(
  letters == "a" ~ "A",
  letters == "b" ~ "B",
  letters == "c" ~ "C",
  TRUE ~ letters
  )

# in case_when case, we only need to specify what to do if condition is met, if not, move onto the next case
# the last line prevents everything being turned to NA if it doesn't match any of the conditions e.g. B as this is already capitalised


# classify students' education level such that...
# students older than 16 are in "sixth form"
# students older than 12 and younger than 16 are in "high school"
# students 12 or younger are in "primary school"

students_big %>% 
  select(ageyears) %>% 
  mutate(education_level = case_when(
    ageyears > 16 ~ "sixth form", # first test
    ageyears > 12 ~ "high school", # if first test failed, try this
    ageyears <= 12 ~ "primary school", # if second test failed, try this
    TRUE ~ nursery # default value, i.e. if all else fails, nursery
  ))

# don't have to worry about more than and less than, as anyone over 16 is going to be dealt with first, so they wont get classed as high school by accident
```


Task - 5 minutes
Using case_when, make a new column estimated_travel_time which is:
Equal to 60 when your region is London.
If you aren’t from London, then:
10 when your travel_to_school is “Walk”.
20 when your travel_to_schoolis “Car”.
For everyone one else then it should equal to 30.

```{r}
students_big %>% 
  select(region, travel_to_school) %>% 
  mutate(estimated_travel_time = case_when(
    region == "London" ~ 60,
    travel_to_school == "Walk" ~ 10, 
     travel_to_school == "Car" ~ 20,
    TRUE ~ 30)) 
```

