---
title: "Recoding"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    css: ../../../styles.css
  pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center')
```

# Learning objectives

1. Understanding what recoding is.
2. Know how to use the functions `if_else`, `recode` and `case_when`.
3. Understand when to use each function.

**Lesson Duration: 1 hour**

In this lesson we will be learning about three functions that help you change observations within a single categorical variable. Changing categorical variables is an essential part of cleaning data. In fact, it's so important it gets a special name: "recoding", and it gets it's own lesson!

Start by loading `tidyverse`, and in this lesson we'll use data from `CodeClanData`.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(CodeClanData)
```

# The `if_else()` function

The function `if_else` lets us do a logical test (just like in `filter`) and then change data depending on the result of that test.

First, lets look at the column `ageyears` from `students_big`.

```{r}
students_big %>% 
  select(ageyears)
```

Let's say we were interested in student ages and wanted to further categorise ages into two groups: those that are age 12 or younger, and those that are older than 12.

We can turn the age column into into a `logical` vector by checking if each element is less than or equal to 12 using `mutate` and a logical operator.

```{r}
students_big %>% 
  select(ageyears) %>% 
  mutate(
    older_than_12 = ageyears > 12
  )
```

All elements that are greater than 12 turn into `TRUE` and all elements to equal to or less  than 12 turn into `FALSE`.

The `if_else()` function lets us transform this vector further in one step - we can pick any values we want to substitute with instead of just `TRUE` and `FALSE`. An example should make this clearer:

```{r}
students_big %>% 
  select(ageyears) %>% 
  mutate(
    older_than_12 = if_else(ageyears > 12, "Older than 12", "12 or younger")
  )
```

The syntax is

<center>
`if_else(boolean expression, value if TRUE, value if FALSE)`

</center>

In many programming languages this is called a **'ternary statement'**, or **'ternary operation'**, because it is composed of **three parts**, as you can see. 

Here are a few more examples.

```{r}
students_big %>% 
  select(gender)

students_big  %>% 
  select(gender) %>% 
  mutate(
    gender = if_else(gender == "M", "Male", "Female")
  )
```

<blockquote class = 'task'>
**Task - 5 minutes**

1. Change the column `handed` to to "Right" and "Left" instead or "R" and "L".
2. In `favourite_physical_activity` change "Football/Soccer" into "Football" and leave all other activities the same.

<details>
<summary> **Solution** </summary>

```{r}
students_big %>% 
  mutate(
    # 1
    handed = if_else(handed == "R", "Right", "Left"),
    # 2
    favourite_physical_activity = if_else(favourite_physical_activity == "Football/Soccer", "Football", favourite_physical_activity)
  )
```

</details>
</blockquote>

# The `recode()` function

The `if_else` function is really useful if you have two values you want to replace. However, what about when you have more? The `recode()` function is another way of recoding values. This particular function allows you to replace values in a vector based on their name.

Here is an example of `recode`. The first argument is the column we want to work with and then we have pairs of *old value = new value*.

```{r}
students_big %>% 
  select(travel_to_school)

students_big %>% 
  select(travel_to_school) %>% 
  mutate(
    travel_to_school = recode(travel_to_school, "Bus" = "bus", "Car" = "car", "Walk" = "walk")
  )
```
You might have noticed that we actually have lots of values for `travel_to_school`.

```{r, message = F}
students_big %>% 
  group_by(travel_to_school) %>% 
  summarise(
    n = n()
  )
```

If we do not specify a new value in `recode` then those levels are unchanged. Or we can set the `.default` argument replace everything that has not already been changed. There is also a `.missing` argument, where you can choose a specific value to replace missing values with.

```{r}
students_big %>% 
  select(travel_to_school) %>% 
  mutate(
    travel_to_school = recode(travel_to_school, "Bus" = "bus", "Car" = "car", "Walk" = "walk", .default = "Other")
  )
```

`recode()` also works on `numeric` and `logical` variables:

```{r}
students_big %>% 
  select(ageyears) %>% 
  mutate(ageyears = recode(ageyears, "12" = 120, .default = 0))
```
Note that despite `ageyears` being a numeric column, the value we want to replace is in quotes. This is because `recode` requires the left side of the equal sign to be in quotation marks.

<blockquote class = "task">

**Task - 10 minutes**

Using the data below, do the following:

* Recode the `subject` numbers to be: 001, 002, 003 and 004
* Recode the `sex` column to be "Male" and "Female"
* Recode the `compl` column to be T = "True", F = "False", and NA otherwise


```{r}
data <- read.table(header = TRUE, text="
 subject sex control cond1 cond2 compl
       1   M     7.9  12.3  10.7  T
       2   F     6.3  10.6  11.1  T
       3   F     9.5  13.1  13.8  N
       4   M    11.5  13.4  12.9  F
")
```

<details>
<summary>**Answers**</summary>
```{r}
data <- data %>% 
  mutate(
    subject = recode(subject, 
                  "1" = "001",
                  "2" = "002",
                  "3" = "003",
                  "4" = "004"
                  ),
    
    sex = recode(sex, 
                  "M" = "Male",
                  "F" = "Female"
                 ), 
    
    compl = recode(compl, 
                  "T" = TRUE,
                  "F" = FALSE,
                  .default = NA
                  )
       )
```

</details>
</blockquote>

# The `case_when()` function

When you want to do more complicated transformations of a vector, you can use the `case_when()` function. 

Using `case_when()` you can write as many cases and outcomes as you like. They have to go on either side of a `~` (tilde). Each time the left hand side is `TRUE`, the vector will have the value on the right. Often we end `case_when()` statements with a `TRUE` on the left-hand-side, that mops up all cases left over.

```{r}
students_big %>% 
  select(ageyears) %>% 
  mutate(
    age_divisible = case_when(
      ageyears %% 2 == 0 ~ "Even",
      ageyears %% 3 == 0 ~ "Divisible by 3",
      ageyears %% 5 == 0 ~ "Divisible by 5",
      TRUE               ~ "Other number"
    )
  )
```

The order in which we do the statements matters. If a value is `TRUE` for two statements, then only the first one will apply.

In the example below, values which are less than or equal to 5, are also less than or equal to 7. However, they are substituted with a value of A, since that comes first. Here we are just using a single vector, but the same thing works inside a data frame.


```{r}
x <- 1:10

case_when(
  x <= 5 ~ "A",
  x <= 7 ~ "B",
  TRUE   ~ "C"
)
```

However, if we swap the order around in the `case_when()` statement, they now all get the value B.

```{r}
x <- 1:10

case_when(
  x <= 7 ~ "B",
  x <= 5 ~ "A",
  TRUE   ~ "C"
)
```

All your substituted values (the values on the right hand side of the tilde `~`) have to be of the same type. The code below will not work, because some of the substituted values are `character` and some are `numeric`. 

```{r, eval = FALSE}
students_big %>% 
  select(ageyears) %>% 
  mutate(
    ageyears = case_when(
      ageyears %% 2 == 0 ~ 2,
      ageyears %% 3 == 0 ~ 3,
      ageyears %% 5 == 0 ~ 5,
      TRUE               ~ "Other number"
    )
  )
```

Be particularly careful if you are returning your original value: if this isn't of the same type as your other substituted values, you'll need to convert it.  

```{r}
students_big %>% 
  select(ageyears) %>% 
  mutate(
    ageyears = case_when(
      ageyears %% 2 == 0 ~ "Even",
      ageyears %% 3 == 0 ~ "Divisible by 3",
      ageyears %% 5 == 0 ~ "Divisible by 5",
      TRUE               ~ as.character(ageyears)
    )
  )
```
<blockquote class = 'task'>

**Task - 5 minutes**

Using `case_when`, make a new column `estimated_travel_time` which is:

* Equal to 60 when your `region` is London. 
* If you aren't from London, then:
  * 10 when your `travel_to_school` is "Walk".
  * 20 when your `travel_to_school`is "Car".
* For everyone one else then it should equal to 30.

<details>
<summary>**Solution**</summary>

```{r}
students_big %>% 
  mutate(
    estimated_travel_time = case_when(
      region == "London" ~ 60,
      travel_to_school == "Walk" ~ 10,
      travel_to_school == "Car" ~ 20,
      TRUE ~ 30
    )
  ) %>% 
  select(estimated_travel_time, region, travel_to_school)
```

</details>
</blockquote>


