---
title: "Assertive programming"
output: html_notebook
---

`assertr` and `stopifnot`

In this lesson you will learn about assertive programming. Assertive programming goes hand-in-hand with testing. When testing, you are checking that your code does what you expect; when carrying out assertive programming, you are **checking your data is in the form you expect.**

# An example with Delhi weather data

## Writing our analysis

```{r}
library(tidyverse)
weather_one <- read_csv("data/delhi_weather_one.csv")
```

And we do some calculations. Let’s find the average temperature, humidity and wind speed. Here we have put the calculations into a function because we’re going to run the calculations again on new data.

```{r}
# below function  named weather_report that returns weather_means and the hot and humid days
weather_report <- function(weather){
  # Our average temperature, humidity and wind speed 
  weather_means <- 
  weather %>% 
  summarise(
    mean_temp = mean(temp),
    mean_humidity = mean(humidity),
    mean_windspeed = mean(wind_speed)
  )

# Show hot and humid days
hot_and_humid <- 
weather %>% 
  filter(temp > 20) %>% 
  filter(humidity > 55)

return(list(weather_means, hot_and_humid))
}

weather_report(weather_one) # call the function on the data we loaded in
```

## New data

Now imagine it’s a month later and we have new weather data. You might just read the data and run your code.

```{r}
weather_two <- read_csv("data/delhi_weather_two.csv")
weather_report(weather_two)
```


Task - 2 minutes

Everything has run without errors, but can you spot something wrong in our output?

For one of the humid days, temp, humidity and wind speed are identical.

# Using `assert`

To avoid this sort of problem we need to use assertive programming. This is the idea that we need to specify what our inputs should be and throw an error if our inputs are unexpected. So far, errors might have seemed like a bad thing to you, but when something is wrong it’s much better to have an error than just getting the wrong results.

To help us do assertive programming we are going to use a package called assertr. This lets us write some “assertions” that describe how we expect the data to look.

```{r}
library(assertr)
```

What do we think are reasonable values for our data? The highest ever temperature in India was 51C, and the lowest ever was -34C, so let’s assume that the temperature will not be higher than 55 or lower than -35. The highest average wind speed average a day in the entire world is 108mph, let’s assume that the wind speed is lower than that. We also need wind speed to be a positive number.

We can use the verify function from assertr to write statements about the data. Anything you can write inside filter you will be able to write inside verify.

```{r}
weather_report <- function(weather){
  
  # put in the checks on the data first !!
  weather %>% 
    verify(temp >= -35 & temp <= 55) %>% 
    verify(wind_speed >= 0 & wind_speed <= 108) %>% 
    # added this
    verify(humidity >= 0 & humidity <=100)
  
  # Our average temperature, humidity and wind speed 
  weather_means <- 
  weather %>% 
  summarise(
    mean_temp = mean(temp),
    mean_humidity = mean(humidity),
    mean_windspeed = mean(wind_speed)
  )

# Show hot and humid days
hot_and_humid <- 
weather %>% 
  filter(temp > 20) %>% 
  filter(humidity > 55)

return(list(weather_means, hot_and_humid))
}
```

Task
Add an appropriate assertion for humidity. Hint: humidity is a percentage.

Let's test it. Now that we’ve added our verify statements, we get an error:

```{r}
weather_report(weather_two)
```

We get the error because the temperature value was unresonably high. Our row also doesn’t meet the assertion for humidity but because of the order we wrote the assertion only the temperature condition is raised. In verify it will stop at the first occurrence of not meeting an assertion and let you know about it. There may be more occurrences in your data but you would need to go in to find them - the point of verify() is to stop the code if the condition is not met, you can then do more work to investigate them.

Task - 5 minutes

Change the order of the assertions so we check humidity before temperature and see how the error message differs.


```{r}
weather_report <- function(weather){
  
  # put in the checks on the data first !!
  weather %>% 
    verify(humidity >= 0 & humidity <=100) %>% 
    verify(temp >= -35 & temp <= 55) %>% 
    verify(wind_speed >= 0 & wind_speed <= 108)
  
  # Our average temperature, humidity and wind speed 
  weather_means <- 
  weather %>% 
  summarise(
    mean_temp = mean(temp),
    mean_humidity = mean(humidity),
    mean_windspeed = mean(wind_speed)
  )

# Show hot and humid days
hot_and_humid <- 
weather %>% 
  filter(temp > 20) %>% 
  filter(humidity > 55)

return(list(weather_means, hot_and_humid))
}

weather_report(weather_two)

```
Error now says assertr stopped execution.

You might have spotted a problem. From looking at the data it’s likely that the wind speed value is also wrong because it was exactly the same as the temperature and humidity (so it’s likely an error in the readings). However our assertions did not manage to catch it because it is within our acceptable range (even though it’s very close to the top end which may also be a sign it’s maybe an error!). Finding numerical errors can be very difficult; sometimes there is almost no way of knowing that you have an error without chasing the data back to the source. However assertive programming can help you avoid extreme errors.


# Using `stopifnot`

Sometimes you will want to do other, more general checks. We can do these using a function from base R called stopifnot. Inside stopifnot we can write a series of statements. If any of the statements are FALSE then we will throw an error. Here is an example.

```{r}
weather_report <- function(weather){
  
  # stop if nrows is not more than 15, or if cols not = 4
  stopifnot(
    nrow(weather) > 15,
    ncol(weather) == 4,
    # Check for any missing values
    drop_na(weather) == weather
  )

  weather %>% 
    verify(temp >= -35 & temp <= 55) %>% 
    verify(wind_speed >= 0 & wind_speed <= 108) %>% 
    verify(humidity >= 0 & humidity <= 100)
  
  # Our average temperature, humidity and wind speed 
  weather_means <- 
  weather %>% 
  summarise(
    mean_temp = mean(temp),
    mean_humidity = mean(humidity),
    mean_windspeed = mean(wind_speed)
  )

# Show hot and humid days
hot_and_humid <- 
weather %>% 
  filter(temp > 20) %>% 
  filter(humidity > 55)

return(list(weather_means, hot_and_humid))
}
```

Check if weather_one meets the assertions we just added.

```{r}
weather_report(weather_one)
```

No errors, so yes it does!





