---
title: "dplyr verbs"
output: html_notebook
---

*Objectives*
- Understand what each 'dplyr' verb does
- Be able to use 'dplyr' verbs
- Be able to handle missing values and nulls while using functions
- Be able to use additional functions that support the 'dplyr' verbs
- Be familiar with the use of the 'magrittr' (pipe) operator

Data wrangling is the process of taking a dataset and manipulate for use in analysis/statistical inference

Dplyr is an R package, part of the tidyverse (groups lots of packages) - the verbs make it all a bit more human friendly: 
summarise, filter, mutate, select

Tibble: rows and columns like a spreadsheet

Select selects columns of interest (e.g. 3 out of 5)
Filter filters out rows (e.g. only people from town X)
Arrange arranges the order of observations (e.g. alphabetical)
Mutate creates a new variable/column
Group by creates groups of observations (e.g. species)
Summarise creates summary statistics 

```{r}
library(tidyverse)
library(CodeClanData)
```
```{r}
tidyverse_logo()
```


Two ways to call functions from libraries:

1. `library(package_name)` followed by `function_name()`
2. `library_name::function_name()`


```{r}
students
```

```{r}
all_deaths
```
Find number of rows in dataset
```{r}
nrow(all_deaths)
```

```{r}
ncol(all_deaths)
```

```{r}
dim(all_deaths)
```

```{r}
names(all_deaths)
```
First 6 rows
```{r}
head(all_deaths)
```

First 10 rows
```{r}
head(all_deaths, 10)
```

Bottom of the data set
```{r}
tail(all_deaths, 2)
```
Glimpse gives us all column names, data types, rows, cols, sample of data
```{r}
glimpse(all_deaths)
```
View shows you a spreadsheet in a separate tab, not usually needed
```{r}
View(all_deaths)
```
Skim shows us overview of missing data and stats of numerical columns
```{r}
skimr::skim(all_deaths)
```


```{r}
select(all_deaths, name, allegiances, gender, nobility, year_of_death)
```

Can remove columns if easier
```{r}
select(all_deaths, -name)
```
```{r}
all_deaths_no_names <- select(all_deaths, -name)
```
Assigning means the result won't show up, unless you put it all in brackets

```{r}
names(all_deaths)
select(all_deaths, book1_GoT, book2_CoK, book3_SoS, book4_FfC, book5_DwD)
select(all_deaths, contains("book"))
select(all_deaths, contains("death"))
`select(all_deaths, 1,2,3,4,5)`
```
#Filter()

For selecting *rows*

`filter` will return all the rows that meet some specified condition

##Logical Operators

`>` greater than
`<` less than
`<=` less than or equal to
`>=` greater than or equal to
`==` check for equality
`%in%` check for equality against *more than one* value
`!=` check for inequality
`!` negate a logical value

```{r}
filter(all_deaths, allegiances == "Lannister")
```
```{r}
filter(all_deaths, allegiances != "Lannister")
```
using %in% allows us to check both lannister strings
```{r}
filter(all_deaths, allegiances %in% c("House Lanniser", "Lannister"))
filter(all_deaths, 
       allegiances == "Lannister",
       allegiances == "House Lannister")
```

```{r}
distinct_alleg <- distinct(all_deaths, allegiances)

arrange(distinct_alleg, allegiances)
```
Filter characters between chapters 5 and 10
```{r}
filter(all_deaths, book_intro_chapter >= 5 & book_intro_chapter <= 10)
```

*Note* AND operations take precedence over OR


1. Find where the year_of_death is less than or equal to 299.
2. Find just the data for the characters “Jon Snow”, “Daenerys Targaryen” and “Samwell Tarly”.
3. Find the females (gender is 0) who are not Lannisters and only return these characters names. 
Hint You will need to combine both of the dplyr verbs we have learned.

```{r}
filter(all_deaths, year_of_death <= 299)
filter(all_deaths, name %in% c("Jon Snow", "Daenerys Targaryen", "Samwell Tarly"))
#select(all_deaths, name (filter(all_deaths, gender == 0, allegiances != "Lannister" !contains("Lannister")))

```

Missed some stuff
```{r}
all_deaths %>%
  mutate(book_of_death = book_of_death * 5)
```
# Converstion functions in mutate()

The conversation functions are
* `as.integer()` # whole number
* `as.double()` # has decimals
* `as.numeric()` # convert to number
* `as.character()` # string
* `as.factor()` # convert to levels
* `as.logical()` # TRUE or FALSE

```{r}
all_deaths %>% 
  mutate(book_of_death = as.character(book_of_death))
```

just cos a number is onscreen doesn't mean its a number that can be operated on

Converting strings of letters to numbers  created NAs
```{r}
all_deaths %>% 
  mutate(name = as.numeric(name))
```

# sum()
```{r}
sum(2,4)
```

```{r}
all_deaths %>% 
  mutate(year_of_death = sum(year_of_death, na.rm = TRUE))
```

# mean()

Returns one value - the mean of the input values

```{r}
mean(c(10,20))
```

```{r}
all_deaths %>% 
  mutate(year_of_death = mean(year_of_death, na.rm = TRUE))
```

Add a new column which is a logical vector, which has TRUE if book_intro_chapter is empty, and FALSE if it isn’t. Hint: you’ll need to use the is.na function.
Change the nobility column to a logical vector.
Find the number of male characters by taking the sum of the gender column (as gender = 1 is male).

```{r}
all_deaths %>% 
  mutate(new_col = is.na(book_intro_chapter))

all_deaths %>% 
  mutate(nobility = as.logical(nobility))

sum(all_deaths$gender)

all_deaths %>% 
  mutate(count_males = sum(gender))
```

# summarise()

Create summary statistics of your data
- this way solves the above task without creating a column you don't need

```{r}

all_deaths %>% 
  summarise(n_names = sum(gender))
```

More common to use group_by with summarise
```{r}
all_deaths %>% 
  # group deaths by Allegiances
  group_by(allegiances) %>% 
  # now count the number of people in each group n() function
  summarise(character_count = n())
```

n function counts number of observations!

Below groups firstly by nobility, then by gender, counts within thins groups
```{r}
all_deaths %>% 
  group_by(nobility, gender) %>% 
  summarise(character_count = n())
```

Task 
Now it’s your turn! Start with the all_deaths data.
If a character doesn’t have a missing value for book_of_death, this means that they have died at some point. Use this to find the allegiances with the highest number of people dead.
Hint You will need to:
Filter the data so we just have characters who have died.
Use group by and summarise to find the number of characters who belonged to each allegiance.
Arrange so the highest values come first.
Try to use a single pipe sequence for this solution.

```{r}
all_deaths %>% 
  filter(!is.na(book_of_death)) %>% 
  group_by(allegiances) %>% 
  summarise(character_count = n()) %>% 
  arrange(desc(character_count))
```

# pull()

e.g. Calculate the average year of death and create a new column which states TRUE if the year of death for the character is above this average, and FALSE if the year of death is below this average.

```{r}
average_year_of_death <-  all_deaths %>% 
  summarise(avg_value = mean(year_of_death, na.rm = TRUE))
```

Now we've calculated the average.. (as a tibble/dataframe)
Below code fails if we don't use it as a single value - can check using class()

```{r}
all_deaths %>% 
  mutate(death_later_than_avg = 
           year_of_death > average_year_of_death)
```
Need to use pull to pull the value out
```{r}
average_year_of_death <-  all_deaths %>% 
  summarise(avg_value = mean(year_of_death, na.rm = TRUE)) %>% 
  pull()
```

now the output is a single value and we can mutate this new column
```{r}
all_deaths %>% 
  mutate(death_later_than_avg = 
           year_of_death > average_year_of_death)
```

If summarising several stats, can include e.g. mean, median within the summarise function

```{r}
state_income_data %>% 
  filter(Index == 'A') %>%
  mutate(income_change = Y2008 - Y2002) %>% 
  summarise(mean = mean(income_change),
            median = median(income_change))
```

