---
title: "Extra `dplyr` Practice - Answers"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    css: ../../../styles.css
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F)
```

# MVP

<br>
The dataset for this minilab is from [Kaggle](https://www.kaggle.com/zynicide/wine-reviews) and contains reviews of different wine varieties (location, winery, price).

<br>

**Question 1.**

Load the tidyverse, read in the data and investigate it, e.g. what are the variable names, variable types, dimensions, how many missing values etc.

<details>
<summary>**Answer**</summary>
```{r, warning = F, message = F}
library(tidyverse)
wine_reviews <- read_csv("data/wine_ratings.csv")

dim(wine_reviews)

names(wine_reviews)

glimpse(wine_reviews)

sum(is.na(wine_reviews))

wine_reviews %>% 
  summarise(across(.cols = everything(), .fns = ~ sum(is.na(.x))))

#view(wine_reviews)
```
</details>

<br>

**Question 2.**

Select columns `title`, `taster_name`, `points`, `price`, and `country`, then **drop all NAs**. Save into a variable called `wine_subset`. Use this variable for all subsequent tasks.

<details>
<summary>**Answer**</summary>
```{r}

wine_subset <- wine_reviews %>% 
  select(title, taster_name, points, price, country) %>% 
  drop_na()

```
</details>

<br>

**Question 3.** 

Find the 5 countries with the highest mean points.

<details>
<summary>**Answer**</summary>
```{r}
wine_subset %>% 
  select(points, country) %>%
  group_by(country) %>% 
  summarise(mean_points = mean(points)) %>% 
  slice_max(mean_points, n = 5)

```
</details>

<br>

**Question 4.** 

Find the 5 countries with the lowest mean prices.

<details>
<summary>**Answer**</summary>
```{r}
wine_subset %>% 
  select(price, country) %>%
  group_by(country) %>% 
  summarise(mean_price= mean(price)) %>% 
  slice_min(mean_price, n = 5)
```
</details>

<br>

**Question 5**

Does the result differ if you use the median instead of the mean for Questions 3 and 4?

<details>
<summary>**Answer**</summary>
```{r}
top_median_points <- wine_subset %>% 
  select(points, country) %>% 
  group_by(country) %>% 
  summarise(median_points = median(points)) %>% 
  slice_max(median_points, n = 5)

top_median_points

lowest_median_price <- wine_subset %>% 
  select(price, country) %>% 
  group_by(country) %>% 
  summarise(median_price = median(price)) %>% 
  slice_min(median_price, n = 5)

lowest_median_price
```

yes - the values and orders of countries are different. Also, when using the median there are more ties, so the code actually results in the top `r nrow(top_median_points)` points and bottom `r nrow(lowest_median_price)` prices unless you add the argument `with_ties = FALSE` to `slice_max` and `slice_min`.
</details>

<br>

**Question 6**

Find the names of all the tasters using `distinct`. Then take a random sample of 5 names using `sample_n`. Save these 5 names into a new variable called `taster_sample` which should be a **character vector** containing only the names. 

<details>
<summary>**Hint**</summary>: You can use the function `pull` to extract a single column as a vector.
</details>

<details>
<summary>**Answer**</summary>
```{r}

taster_sample <- wine_subset %>% 
  distinct(taster_name) %>% 
  sample_n(5) %>% 
  pull()

taster_sample
```
</details>

<br>

**Question 7**

Do the same thing for `country`, i.e. create a **character vector** called `country_sample` using `distinct` and `pull`, but this time use `sample_frac` to pick 10 countries at random. 

<details>
<summary>**Hint**</summary>: What is the proportion of countries you want? $\frac{Part}{Whole}$
</details>

<details>
<summary>**Answer**</summary>
```{r}

country_sample <- wine_subset %>% 
  distinct(country) %>% 
  sample_frac(10/n()) %>% # can replace with 42 or with nrow(.)
  pull()

country_sample
```
</details>

<br>

**Question 8**

Filter the rows of `wine_subset` so you are left only with data from the 5 tasters in your `taster_sample` and from the countries in your `country_sample`. Save this into a new variable called `wine_sample`.

<details>
<summary>**Answer**</summary>
```{r}
wine_sample <- wine_subset %>% 
  filter(taster_name %in% taster_sample,
         country %in% country_sample)

wine_sample
```
</details>

<br>

**Question 9**

Using your `wine_sample`, create a new column `difference_from_avg` which contains the difference between each wine's rating (`points`) and the average rating for that taster, then use arrange to see the highest positive difference first in the column. [**Hint:** you will first have to calculate the average points for each taster]

<details>
<summary>**Answer**</summary>
```{r}

wine_sample %>% 
  group_by(taster_name) %>% 
  mutate(avg_rating = mean(points)) %>% 
  mutate(difference_from_avg = points - avg_rating) %>% 
  arrange(desc(difference_from_avg))

```
</details>
