---
title: "errors"
output: html_notebook
---

Here to learn some common sources of errors and techniques to solve these

-Watch out for spelling errors
tabs help you autofill to ensure you get it bang on
can use base R rm() function to remove a variable if it's hanging around unwanted, but restart the R session will work too

fcuk package is a fun one that anticipates typos and makes suggestions

-Upstream change
Made a change but didn't update rest of code to get onboard, can use RNotebook to run chunks above or below to deal with this

-Conflict of functions
Due to  different functions with the same name in different packages

```{r}
library(tidyverse)
library(MASS)
```
Warning shows us that selected is masked from dplyr, mass::select will have dominance as MASS loaded last

```{r}
mrcars %>% 
  select(mpg)
# see? masked! can use R help to see what the mass::select does

# can do this to specify
mtcars %>% 
  dplyr::select(mpg)
```

can use conflicts() to see conflicts

```{r}
variable <- c(1,2,3)
```
In the console, this input is still needing finished (see plus signs in console). It's open and will cause an error.


`no package called x`
Maybe you loaded the library in wrong, or typo in the package loading (e.g. tidyverse is a package of packages)


# Other tips 
- Build code up, line by line until you find the error
- Read the error message, can point you in right direction
- google the error message
- Step away from your computer
- Rubber duck debugging (talk it through!)
- Get someone to look at your code
- Warnings aren't errors, but useful still - e.g. comparing vector c(1,2,3,4,5,6,7,8,9,10) == c(2,3,4) 