---
title: "Joins lab"
output: html_notebook
---

```{r}
library(tidyverse)

actors <- read_csv("data/actors.csv")
movies <- read_csv("data/movies.csv")
roles <- read_csv("data/roles.csv")

```

```{r}
movies %>% 
  filter(name == "Little Mermaid, The")
```

id is 192017

```{r}
little_mermaid_roles <- roles %>% 
  filter(movie_id == 192017)
```

```{r}
little_mermaid_roles %>% 
  left_join(actors, by = c("actor_id" = "id"))
little_mermaid_roles
```

```{r}
little_mermaid_roles %>% 
  inner_join(actors, by = c("actor_id" = "id"))
little_mermaid_roles
```

Data is the same, meaning there are no columns in the left table that have missing data when merged with the right using id.

```{r}
actors_more_than_once <- actors %>% 
  filter(film_count > 1)
actors_more_than_once

actors_more_than_once <- actors_more_than_once %>% 
  left_join(roles, by = c("id" = "actor_id"))

actors_more_than_once
```


```{r}
actors_more_than_once <- actors_more_than_once %>% 
  left_join(movies, by = c("movie_id" = "id"))

actors_more_than_once
```
For each actor, find the average rating of the films they appeared. Find the five actors with the highest average rank of films they appeared in.
```{r}
actors_more_than_once2 <- actors_more_than_once %>% 
  group_by(first_name, last_name) %>% 
  mutate(rank = as.numeric(rank)) %>% 
  summarise(average_rank = mean(rank, na.rm = TRUE)) %>% 
  select(first_name, last_name, average_rank) %>% 
  ungroup %>% 
  slice_max(average_rank, n = 5)


actors_more_than_once2
```

Find the actor who has appeared in most films. List the films they appeared in, the year the film was released and their role in the film.

```{r}
actors %>% 
  arrange(desc(film_count))
```
Kevin Bacon
```{r}
actors %>% 
  filter(first_name == "Kevin")
```
His ID is 22591

```{r}
movies %>% 
  left_join(roles, by = c("id" = "movie_id")) %>% 
  filter(actor_id == 22591) %>% 
  select(name, year, role)
```










