---
title: "Unit testing"
output: html_notebook
---

Create a function to test

```{r}
is_leap_year <- function(year){
  
  if (year %% 100 == 0){
    if (year %% 400 == 0){
      return(TRUE)
    } else{
      return(FALSE)
    }
  } 
    
  if (year %% 4 == 0){
    return(TRUE)
  } else {
    return(FALSE)
  }
  
}
```


Task - 5 minutes

Write out what this function is doing, and why it is written in this format (e.g. Why do we test the 100 and 400 years first?)

Function returns true if the year input is divisible by 100 and 400. Then checks if it's divisble by 4 for another return TRUE. Have to catch the larger numbers first as many things are divisible by 4 and would therefore return TRUE.

## `testthat` package

testthat tries to make testing as fun as possible, so that you get a visceral satisfaction from writing tests. Testing should be addictive, so you do it all the time.


```{r}
library(testthat)
```

expect_equal checks that the objects give you an answer you expect
expect_error checks that code throws an error

e.g.
```{r}
a <- 10
# we expect a to be 10, if it does - carry on, no error
expect_equal(a,10)
expect_equal(a,11)
```

### Tests that pass

The workflow of unit testing is best described by example. Let’s write a test that checks if real leap years return a true value from our function.


# first test: two expectations
```{r}
# first test: two expectations
test_that("Real leap years return true", {
  expect_equal(is_leap_year(2020), TRUE)
  expect_equal(is_leap_year(2000), TRUE)
})
```

Task - 5 minutes

Take the following test code and write down an explanation of all the different parts:

What is the test name?
"Character input returns an error"
What is the expectation?
that "year" will return an error
What are you expecting this test to do?
Pass
Think carefully about the logic of this one before you write down your assumptions… you may want to try and run it to see what happens first.

```{r}
test_that("Character input returns an error", {
  expect_error(is_leap_year("year"))
})
```

### Tests that fail

```{r}
test_that("Non-numeric input returns an error", {
  expect_error(is_leap_year("year"))
  expect_error(is_leap_year(TRUE))
})
```
Task - 5 minutes

Which of our tests has failed?
2nd test
Why has it failed?
no error when one was expected
Why has our other test passed?
because an error was expected 


So now let’s change the function code to make this test pass:

```{r}
is_leap_year <- function(year){

  # add a condition, that makes sure our function ONLY takes in numeric input
  if (!is.numeric(year)){
    stop("Non-numeric input 😿")
  }
    
  if (year %% 100 == 0){
    if (year %% 400 == 0){
      return(TRUE)
    } else{
      return(FALSE)
    }
  } 
    
  if (year %% 4 == 0){
    return(TRUE)
  } else {
    return(FALSE)
  }
  
}
```

Now our function has been updated so that the first thing it will do is check for non-numeric input. That should allow our test to work now:

```{r}
test_that("Non-numeric input returns an error", {
  expect_error(is_leap_year("year"))
  expect_error(is_leap_year(TRUE))
})
```

Woo test passed!

Task - 5 minutes

Write a test that checks whether 2017 and 1900 are correctly identified as not being leap years by our is_leap_year() function.

```{r}
test_that("2017 and 1900 correctly identified", {
  expect_false(is_leap_year(2017))
  expect_false(is_leap_year(1900))
})
```

Could have used expect_equal here, there's also expect_type, expect_gt (greater than), expect_match (for pattern matching)


