---
title: "coding best practice"
output: html_notebook
---

Code style is about much more than looking good
Most important person to think about is future you - will you be understand your code?

We adopt the tidyverse style guide. We've summarised the important parts of the style guide that apply to you just now.

- lower-case letters, numbers, and underscores - `snake_case`
others include PascalCase, camelCase, skewer-case, SCREAMING_SNAKE_CASE, nocase

- avoid df, data, test etc. as they are non descriptive

- objects should be nouns and functions should be verbs
Think of it like "this function will... `convert_miles_to_km`

- try not to overwrite functions and objects already in R
leads to conflicts

e.g.'s
1. BabyWeightData - bad only cos of Pascal, but otherwise good
2. data - too vague
3. baby_weight_data - v good
4. babies - too vague but still good
5. weighting_babies - naming an object with a verb

# Internal structure
 - always start with your library calls and then reading in data
 - use empty space to split code into sections
 - use commented lines of _ to show larger sections
 
 e.g. 
 library(x)
 library(y)
 
 # doing thing one -----------------------
 
 code
 
 # doing thing two -----------------------
 
# Spacing and assignment 
 
 - put a space after a comma
 - do not put spaces directly inside or outside parentheses
 - use <- for assignment, and put spaces either side
 - also use spaces on either side of = and ==
 - extra space are okay if they help align <- or =
 - there is also an R Studio shortcut Ctrl +i to help with alignment
 
='s are everywhere! <- lets us know exactly when assignments are happening 

# Line length

- Code shouldn't be more than 80 characters
- Can add margin to prevent going over 80
- Can split function across multiple lines e.g. 
tibble(
  x = x,
  y = y,
)
- Multiple line function calls should have the function name and opening bracket on one line, arguments on next lines one tab length in, and finish with the closing bracket on a single line (RStudio does a lot of this for you)
- If writing in a notebook your code will wrap at the end of the line, but it's best to avoid this happening by keeping lines short
- Break code up using pipes, this can stop lines from getting too long

 
 e.g. big_function(
  something = "that",
  requires = many,
  arguments = "some of which may be long"
)

# Quotes

- use double rather than single quotes

```{r}
"string"
```
```{r}
'string'
# R puts it back in doubles, R has a preference for that
'dont\'t use single quotes for strings'
```

# Comments

Handy for explaining processes.

RNotebook is good for writing text outside of code chunks

```{r}
# can use hash inside code chunk
then_we_can_write_our_code <- 1

```

Ideally want code to be understandable, code is there to explain the human reasoning, results of analysis etc

Don't explain every step, or comment when a section of notebook would be better

# What should make it into the final code?

Want the minimum amount of code that gets where you want to go. Don't need to show code of understanding data.

Remember that your initial code may get cleaned up at the end and all the extra gubbins removed.

Styler add in can easily tidy up code for tidyverse style


# Other best practices

- be reproducible - which is one of the main advantages of a scripting language as you can see exactly how the analysis was performed - unlike excel
- short is better than long
- fast code is better than slow, but not really a problem unless silly amounts of code, don't worry
- avoid premature optimisation, profiling code is when you see how long each bit takes and trying to speed it up - but don't worry too much
- don't be afraid to rewrite to make code clearer or faster
- but clarity is better than conciseness




 